Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-27 sections/eval.tex
Update on Overleaf.

Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. SpGEMM focuses on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
The inner product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. 
Inner product involves multiplying two matrices by taking dot products of corresponding rows and columns.
Outer product computes the product of all pairs of elements from the input matrices.
More detailed explanation is available in Section 2.2 of ~\cite{} \changwan{gamma}.
\changwan{Inner, asym bad, }

Figure~\ref{spgemm_small} 
TACO does not support outer product due to its lack of support for scatter into a 2D workspace.

@@

@@

\subsubsection{The Benefits of Kernel Fusion}

>>>>

Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. SpGEMM focuses on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
The inner product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. 
Inner product involves multiplying two matrices by taking dot products of corresponding rows and columns.
Outer product computes the product of all pairs of elements from the input matrices.
More detailed explanation is available in Section 2.2 of ~\cite{zhang2021gamma}.
Figures ~\ref{spgemm_small} and ~\reef{spgemm_large} show

TACO does not support outer product due to its lack of support for scatter into a 2D workspace.

@@

\begin{table}[htbp]
    \centering
    \caption{SpMV Sample Matrices}
    \label{spmv_sample_matrices}
    \begin{tabular}{|l|c|c|c|c|}
     \textbf{Spy} &
     \includegraphics[width=3cm]{spmv_matrices/saylr4.png} &
     \includegraphics[width=3cm]{spmv_matrices/heart3.png} &
     \includegraphics[width=3cm]{spmv_matrices/toeplitz_large_band.png} &

    \end{tabular}
\end{table}

@@

\subsubsection{Evaluating Finch Logic}
To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) multiple elementwise operations. Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined execution and even takes advantage of vector instructions which are not yet present in Finch.

### 2024-03-27 sections/lang.tex
Update on Overleaf.

\begin{figure}
\noindent % This ensures the minipages fill the page width
\begin{minipage}{0.4\linewidth}
\begin{align*}
    \finchliteral(val \in \mathbb{V}) &:= \text{\mintinline{julia}{val}} \\
    \finchvalue(ex \in \mathbb{S}, type \in \mathbb{T}) &:= \text{\mintinline{julia}{ex :: type}} \\
    \finchtensor(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
    \finchindex(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
    \finchvar(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
    \finchextent(a \in E, b \in E) &:= \text{\mintinline{julia}{a : b}} \\
    \finchcall(f \in E, args\ldots \in E) &:= \text{\mintinline{julia}{f(args...)}} \\
    \finchaccess(tns \in T, idxs\ldots \in E) &:= \text{\mintinline{julia}{tns[idxs...]}} \\
    \finchfreeze(tns \in T) &:= \text{\mintinline{julia}{@freeze(tns)}} \\
    \finchthaw(tns \in T) &:= \text{\mintinline{julia}{@thaw(tns)}} \\
    \finchmode(tns \in T) &:= \text{\mintinline{julia}{@mode(tns)}} \\
\end{align*}
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{align*}
    \mathbb{V} &:= \text{the set of all values.} \\
    \mathbb{S} &:= \text{the set of all symbols.} \\
    \mathbb{T} &:= \text{the set of all types.} \\
    I &:= \finchindex \\
    A &:= \finchaccess \\
    V &:= \finchvar \\
    T &:= \finchtensor \\
    E &:= \finchliteral \mid \finchvalue \mid \finchindex \\
        &\quad \mid \finchvar \mid \finchextent \mid \finchcall \mid \finchaccess \\
    S &:= \finchassign \mid \finchloop \mid \finchdefine \\
        &\quad \mid \finchsieve \mid \finchblock \\
\end{align*}
\end{minipage}%
\begin{align*}
    \finchdeclare(tns \in T, init \in E, dims\ldots \in E) &:= \text{\mintinline{julia}{tns .= init(dims...)}} \\
    \finchassign(lhs \in A, op \in E, rhs \in E) &:= \text{\mintinline{julia}{lhs \texttt{<<op>>=} rhs}} \\
    \finchloop(idx \in I, range \in E, body \in S) &:= \text{\mintinline{julia}{for idx = range; body end}} \\
    \finchdefine(var \in V, val \in E, body \in S) &:= \text{\mintinline{julia}{let var = val; body end}} \\
    \finchsieve(cond \in E, body \in S) &:= \text{\mintinline{julia}{if cond; body end}} \\
    \finchblock(bodies\ldots \in S) &:= \text{\mintinline{julia}{begin; bodies... end}}
\end{align*}
\caption{The syntax of the finch language. Note that instead of a classic EBNF
form, we have constrained the classes of subterms on the left hand side to allow
for descriptive names of the arguments. Compare this grammar to the Concrete
Index Notation of TACO \cite[Figure~3]{kjolstad_tensor_2019}, noting the
addition of multiple left hand sides through code blocks, and the explicit
initialization and finalization of tensors.}\label{fig:syntax}
\end{figure}

@@

We break our description of the semantics of Finch into three parts, those
relating to lowering the language itself in Figure~\ref{fig:semantics_core},
those pertaining to the lifecycle of tensors in
Figure~\ref{fig:semantics_lifecycle}, and the semantics of the looplets in
Figure~\ref{fig:semantics_looplets}. Note that in addition to the rules seen
here, we also apply extensive symbolic simplification of the program whenever we
can. In the implementaiton of the language we have heuristics for when to run a
simplification pass, but for the purposes of understanding the semantics of the
language we should understand simplification rules as occuring whenever possible
before continuing to execute the program as normal.

@@

\begin{figure}
    \centering
    \small

@@

    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow true}
        \infer1[$SieveTrue$]{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow false}
        \infer1[$SieveFalse$]{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
    \end{prooftree}

    \caption{Basic evaluation semantics, roughly defining most of these language
    constructs to function similarly to their classical definitions. Note that
    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
    environment and another state $t$ corresponding to the state in the
    host language. Several Looplets introduce variables into the embedding
    language, which may be read when evaluating the \finchvalue node.}
    \label{fig:semantics_core}
\end{figure}

\begin{figure}
    \centering
    \small

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(tns) = tns'}
    \hypo{\langle init, s \rangle \Rightarrow init'}
    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
    \infer4[$Declare$]{\langle\finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle \finchdeclare(tns', init', dims'...), t \rangle)}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = update}
    \hypo{e(tns) = tns'}
    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
    \end{prooftree}
    \vspace{12pt}

@@

    \caption{Lifecycle evaluation semantics. The state $s$ of the finch compiler
    is a tuple $(e, t)$ of a variable value environment and another state $t$
    corresponding to the state in the host language. All of the lifecycle
    functions are designed to be implemented and executed in the host language,
    but these semantics enforce that each of these functions may update state in
    the host language and flip the mode of the tensor between read and update.}
    \label{fig:semantics_lifecycle}
\end{figure}

\begin{figure}
    \centering
    \small

@@

    \caption{Looplet evaluation semantics. 
    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
    environment and another state $t$ corresponding to the state in the host
    language.  Note that $E$ is an evaluation context that applies anywhere in
    the syntax tree. The nonlocal evaluations of Looplets are what allow looplets to
    hoist conditions and subranges out of loops. However, this also means we must specify
    the priority in which we apply looplet rules, which is as follows:
    $Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup$
    Many looplets, most notably the thunk looplet, introduce variables into the
    host language environment.  While variables introduced by a looplet may be
    modified by the looplet itself (steppers often increment some state
    variables), we forbid child looplets from modifying any state variables that
    they didn't introduce. This allows us to treat the \finchvalue node as a
    constant.
    %
    Note: the $Simplify$ rule references $algebra$, which is our variable
    defining a set of straightforward simplification rules. These rules include
    simple properties like $x * 0 \rightarrow 0$ to more complicated ones such
    as constant propagation. We omit the full set of rules for brevity, but
    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}
    \label{fig:semantics_looplets}
\end{figure}

>>>>

\begin{figure}
\scriptsize
\noindent % This ensures the minipages fill the page width
\begin{minipage}{0.4\linewidth}
\begin{align*}
    \finchliteral(val \in \mathbb{V}) &:= \text{\mintinline{julia}{val}} \\
    \finchvalue(ex \in \mathbb{S}, type \in \mathbb{T}) &:= \text{\mintinline{julia}{ex :: type}} \\
    \finchtensor(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
    \finchindex(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
    \finchvar(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
    \finchextent(a \in E, b \in E) &:= \text{\mintinline{julia}{a : b}} \\
    \finchcall(f \in E, args\ldots \in E) &:= \text{\mintinline{julia}{f(args...)}} \\
    \finchaccess(tns \in T, idxs\ldots \in E) &:= \text{\mintinline{julia}{tns[idxs...]}} \\
    \finchfreeze(tns \in T) &:= \text{\mintinline{julia}{@freeze(tns)}} \\
    \finchthaw(tns \in T) &:= \text{\mintinline{julia}{@thaw(tns)}} \\
    \finchmode(tns \in T) &:= \text{\mintinline{julia}{@mode(tns)}} \\
\end{align*}
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{align*}
    \mathbb{V} &:= \text{the set of all values.} \\
    \mathbb{S} &:= \text{the set of all symbols.} \\
    \mathbb{T} &:= \text{the set of all types.} \\
    I &:= \finchindex \\
    A &:= \finchaccess \\
    V &:= \finchvar \\
    T &:= \finchtensor \\
    E &:= \finchliteral \mid \finchvalue \mid \finchindex \\
        &\quad \mid \finchvar \mid \finchextent \mid \finchcall \mid \finchaccess \\
    S &:= \finchassign \mid \finchloop \mid \finchdefine \\
        &\quad \mid \finchsieve \mid \finchblock \\
\end{align*}
\end{minipage}%
\begin{align*}
    \finchdeclare(tns \in T, init \in E, dims\ldots \in E) &:= \text{\mintinline{julia}{tns .= init(dims...)}} \\
    \finchassign(lhs \in A, op \in E, rhs \in E) &:= \text{\mintinline{julia}{lhs <<op>>= rhs}} \\
    \finchloop(idx \in I, range \in E, body \in S) &:= \text{\mintinline{julia}{for idx = range; body end}} \\
    \finchdefine(var \in V, val \in E, body \in S) &:= \text{\mintinline{julia}{let var = val; body end}} \\
    \finchsieve(cond \in E, body \in S) &:= \text{\mintinline{julia}{if cond; body end}} \\
    \finchblock(bodies\ldots \in S) &:= \text{\mintinline{julia}{begin; bodies... end}}
\end{align*}
\caption{The syntax of the finch language. Note that instead of a classic EBNF
form, we have constrained the classes of subterms on the left hand side to allow
for descriptive names of the arguments. Compare this grammar to the Concrete
Index Notation of TACO \cite[Figure~3]{kjolstad_tensor_2019}, noting the
addition of multiple left hand sides through code blocks, and the explicit
initialization and finalization of tensors.}\label{fig:syntax}
\end{figure}

@@

We present a sketch of a small-step operational semantics, showing how to execute a Finch program in a host language.
We break our description of the semantics of Finch into three parts, those
relating to lowering the language itself in Figure~\ref{fig:semantics_core},
those pertaining to the lifecycle of tensors in
Figure~\ref{fig:semantics_lifecycle}, and the semantics of the looplets in
Figure~\ref{fig:semantics_looplets}. Note that in addition to the rules seen
here, we also apply extensive symbolic simplification of the program whenever we
can. In the implementaiton of the language we have heuristics for when to run a
simplification pass, but for the purposes of understanding the semantics of the
language we should understand simplification rules as occuring whenever possible
before continuing to execute the program as normal.

@@

%What are values? (objects in the host language)
%What is t really? (

\begin{figure}
    \centering
    \footnotesize

@@

    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow true}
        \infer1[$SieveTrue$]{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}
    \end{prooftree}
    \hfill
    \begin{prooftree}
        \hypo{\langle cond, s \rangle \Rightarrow false}
        \infer1[$SieveFalse$]{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(tns) = tns'}
    \hypo{\langle init, s \rangle \Rightarrow init'}
    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
    \hypo{\langle declare(tns', init', dims'...), t \rangle \rightarrow t'}
    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], t')}
    \end{prooftree}
    \vspace{12pt}

    \begin{prooftree}  
    \hypo{s = (e, t)}
    \hypo{e(\finchmode(tns)) = update}
    \hypo{e(tns) = tns'}
    \hypo{\langle freeze(tns'), t \rangle \rightarrow t'}
    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], t')}
    \end{prooftree}
    \vspace{12pt}

@@

    \caption{Basic evaluation semantics, roughly defining most of these language
    constructs to function similarly to their classical definitions. Note that
    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
    environment and another state $t$ corresponding to the state in the
    host language. Several Looplets introduce variables into the embedding
    language, which may be read when evaluating the \finchvalue node.
    %
    The state $s$ of the finch compiler
    is a tuple $(e, t)$ of a variable value environment and another state $t$
    corresponding to the state in the host language.
    This means that rules which modify $t$ are running in the host language. All of the lifecycle
    functions are designed to be implemented and executed in the host language,
    but these semantics enforce that each of these functions may update state in
    the host language and flip the mode of the tensor between read and update}
    \label{fig:semantics_core}
\end{figure}

\begin{figure}
    \centering
    \footnotesize

@@

    \caption{Looplet evaluation semantics. 
    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
    environment and another state $t$ corresponding to the state in the host
    language. This means that rules which modify $t$ are running in the host language. Note that $E$ is an evaluation context that applies anywhere in
    the syntax tree. The nonlocal evaluations of Looplets are what allow looplets to
    hoist conditions and subranges out of loops. However, this also means we must specify
    the priority in which we apply looplet rules, which is as follows:
    $Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup$
    Many looplets, most notably the thunk looplet, introduce variables into the
    host language environment.  While variables introduced by a looplet may be
    modified by the looplet itself (steppers often increment some state
    variables), we forbid child looplets from modifying any state variables that
    they didn't introduce. This allows us to treat the \finchvalue node as a
    constant.
    %
    Note: the $Simplify$ rule references $algebra$, which is our variable
    defining a set of straightforward simplification rules. These rules include
    simple properties like $x * 0 \rightarrow 0$ to more complicated ones such
    as constant propagation. We omit the full set of rules for brevity, but
    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}
    \label{fig:semantics_looplets}
\end{figure}
